### 2017-11-09 vipassana-for-hackers.tex
cleaned images, finished references

Well-meaning Neuroscience papers which completely miss the mechanics of nervous-system-based meditation techniques and their core outcomes.
Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. [1]
Descriptions of Vipassana from an individual and subjective perspective, usually given by someone who has attended one course. (``Oh my god! It was sooooo intense!'')

@@

This is not a comparison with other meditation techniques, though I will address other techniques for context. It is worth noting that there are a few meditation techniques labelled “Vipassana” and this paper deals specifically with Vipassana as taught by S.N. Goenka.

@@

A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' [2], as per Theravada. I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings. However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.

@@

The fundamental difficulty in any description of Vipassana is this: It is impossible to convey a description of Vipassana in terms known to someone who has not experienced it. During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother [3]. Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.

@@

Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' [4] and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely. LISP has some very interesting properties: Homoiconicity, metaprogramming, self-mutation and self-replication, recursion and trampolining, and — in the case of modern LISPs and other languages with strong concurrency models — hardened notions of immutability and the reification of time.

@@

\begin{center}
  \LARGE{1. Meditation is a thing}
\end{center}

@@

\begin{figure}[h]
  \includegraphics[width=\linewidth]{meditation-spectrum-1.jpeg}
  \caption{Sample meditation spectrum.}
  \label{fig:meditation-spectrum}
\end{figure}

For the purposes of this paper, we will keep all notions of “mindfulness” in the “meditative” camp, on the right. Mindfulness is not meditation. This paper will be discussing only two practices from this list: Anapana and Vipassana. An illustrated definition of “awareness” and “attention” follows, in the section entitled “On Attention”.

2. Your brain and nervous system have a tight feedback loop

When we say “tight feedback loop”, this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.

@@

3. Dissolution of the body
is a thing

@@

To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like this:

As the field of sensation progresses to the left, the sensations felt become increasingly difficult to describe. “Surreal hot / cold” refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid. “Spatial distortion” refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception. This may include reorientation in space (“I was upside-down!”) or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.

“Throbbing” refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a “vibration”. Beyond vibration is the “invisible”, where a particular body part is no longer seen and a deeper part of the body is experienced directly. The first such example in most cases tends to be the skin, since the external surface of the body is the first part of the nervous system the course instructions ask the meditator to focus on. Beyond this the meditator may find muscle tissue, bones, the cardiovascular system, or something else entirely. Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so “something else entirely” is likely to be the most common case.

The sample spectrum above is missing many, many sensations. I also feel it necessary to call out that the spectrum is not serial; although I have felt “vibrations” and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum. However, the diagram hopefully provides a reference point for the type of readily-accessible sensations which Vipassana begins with (on the right) and the types of sensations (or absence of sensation) which will convince a person that complete dissolution of the body not only possible but an entirely natural process.

Virtual Reality
and
The Matrix Paradox

@@

This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at Facebook F8 in 2015 [5]. I'm using this slide because it's nicely laid out and because it is incomplete.

Abrash discusses the nature of the ``Matrix Paradox'' [6] as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. The following might be a slightly better model of reality:

Though a classic sense diagram such as Mr. Abrash’s might describe proprioception [7, 8], nocioception [9], and hunger [10] as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain. Touch is strictly an external sense dealing with skin. These are internal senses.

@@

On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' [11]) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience. On the topic of lacking proprioception: “We have words like ‘deaf’ and ‘blind’. We do not have words for being deaf or blind to one’s own body.” In the Matrix, perhaps this is either a bug or a feature flag.

Now that we have our diagram for visualizing the use of our senses, we can address them in terms of various altered mental states. Before we get there, however, we will try to come up with a meaningful visual definition of “attention” or “awareness” with respect to our senses.

@@

The funny thing about Vipassana is that you are doing it already... you're just terrible at it. Vipassana is the inspection of one's own nervous system with narrowly-focused attention but the precision of this focus is not prescribed by the instruction given for the technique. “As narrow as possible”, perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip. Your attention is, in part, already on your nervous system, with some degree of focus. To the extent that it is, you are doing Vipassana.

@@

On the first day, your attention might look something like this:
…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent “outside” the body on thoughts and emotions. By the fourth day, your attention will often take a shape like this instead:

@@

If you can’t feel pain in your knees despite the fact that the pain is “real”, you have effectively eliminated (some) sensory input. This is possible for all sensory input. The 5 classic sense-doors will quite clearly shut down in deep meditative states. Just because your eyes are closed does not make you blind. Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness. This blob which I ``see'' disappears in deep meditative states. Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
Blue Bars:
The New Distractions

@@

These ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar. These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana. Everything in blue, unfortunately, falls into the “explaining the concept of ‘white’ to a birth-blind man” category of unexplainable sensory input.

@@

This statement is meant operationally, not existentially. We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event — every operation — occurring in the nervous system has a corresponding event in the brain, and so forth. This mutually recursive process will eventually wear itself out. If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives. However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is “argh, I hate pain”, which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.

@@

Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body. From ``Where Am I?'' [11]:

“The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.”

“Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.”

@@

There is a sketch in ``Where Am I?'' [11] about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:

“Honey, don’t forget what the half-life is on the autonomic nervous system.”

@@

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide is from his 2013 talk ``Performance Testing Java Applications'' [12] and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.

@@

[1] ``The Newsroom'':
\url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
[2] Five Precepts:
\url{https://en.wikipedia.org/wiki/Five_Precepts}
[3] S.N. Goenka -- Experience the Truth
\url{https://www.youtube.com/watch?v=mcbGbN8pFyk&list=PLlb2aTuor6J9j8hACeX6GUmd4WF7QJsbu&index=2}
[4] Eric S. Raymond -- How To Become A Hacker
\url{http://www.catb.org/esr/faqs/hacker-howto.html}
[5] Michael Abrash -- Why Virtual Reality Will Matter to You
\url{https://www.youtube.com/watch?v=UDu-cnXI8E8}
[6] Simulation Hypothesis
\url{https://en.wikipedia.org/wiki/Simulation_hypothesis}
[7] The use of visual feedback in restoring brain function:
\url{https://academic.oup.com/brain/article/132/7/1693/328686}
[8] Proprioception:
\url{https://en.wikipedia.org/wiki/Proprioception}
[9] Nocioception:
\url{https://en.wikipedia.org/wiki/Nociception}
[10] Brain activity in hunger and satiety: an exploratory visually stimulated FMRI study
\url{https://www.ncbi.nlm.nih.gov/pubmed/18292747}
[11] RadioLab -- Where Am I?
\url{http://www.radiolab.org/story/91524-where-am-i/}
[12] Martin Thompson -- Performance Testing Java Applications
\url{https://www.youtube.com/watch?v=1DuMvpwWHH4}

>>>>

Well-meaning Neuroscience papers which completely miss the mechanics of nervous-system-based meditation techniques and their core outcomes.
Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. \cite{thenewsroom}
Descriptions of Vipassana from an individual and subjective perspective, usually given by someone who has attended one course. (``Oh my god! It was sooooo intense!'')

@@

This is not a comparison with other meditation techniques, though I will address other techniques for context. It is worth noting that there are a few meditation techniques labelled ``Vipassana'' and this paper deals specifically with Vipassana as taught by S.N. Goenka.

@@

A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' \cite{fiveprecepts}, as per Theravada. I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings. However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.

@@

The fundamental difficulty in any description of Vipassana is this: It is impossible to convey a description of Vipassana in terms known to someone who has not experienced it. During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother. \cite{experiencethetruth} Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.

@@

Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely. LISP has some very interesting properties: Homoiconicity, metaprogramming, self-mutation and self-replication, recursion and trampolining, and — in the case of modern LISPs and other languages with strong concurrency models — hardened notions of immutability and the reification of time.

@@

\vspace{1cm}
\begin{center}
  \LARGE{1. Meditation is a thing}
\end{center}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{images/meditation-spectrum-1.jpeg}
  \caption{Sample meditation/meditative spectrum.}
  \label{fig:meditation-spectrum-1}
\end{figure}

For the purposes of this paper, we will keep all notions of ``mindfulness'' in the ``meditative'' camp, on the right. Mindfulness is not meditation. This paper will be discussing only two practices from this list: Anapana and Vipassana. An illustrated definition of ``awareness'' and ``attention'' follows, in the section entitled ``On Attention''.

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{images/meditation-spectrum-3.jpeg}
  \caption{Sample meditation spectrum.}
  \label{fig:meditation-spectrum-3}
\end{figure}

\vspace{1cm}
\begin{center}
  \LARGE{2. Your brain and nervous system have}\\
  \LARGE{a tight feedback loop}
\end{center}

When we say ``tight feedback loop'', this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.

@@

\vspace{1cm}
\begin{center}
  \LARGE{3. Dissolution of the body is a thing}
\end{center}

@@

To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like the figure on the following page.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/sensation-spectrum.jpeg}
  \caption{Sample sensation spectrum.}
  \label{fig:sensation-spectrum}
\end{figure}

As the field of sensation progresses to the left, the sensations felt become increasingly difficult to describe. ``Surreal hot / cold'' refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid. ``Spatial distortion'' refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception. This may include reorientation in space (``I was upside-down!'') or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.

``Throbbing'' refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a ``vibration''. Beyond vibration is the ``invisible'', where a particular body part is no longer seen and a deeper part of the body is experienced directly. The first such example in most cases tends to be the skin, since the external surface of the body is the first part of the nervous system the course instructions ask the meditator to focus on. Beyond this the meditator may find muscle tissue, bones, the cardiovascular system, or something else entirely. Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so ``something else entirely'' is likely to be the most common case.

The sample spectrum above is missing many, many sensations. I also feel it necessary to call out that the spectrum is not serial; although I have felt ``vibrations'' and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum. However, the diagram hopefully provides a reference point for the type of readily-accessible sensations which Vipassana begins with (on the right) and the types of sensations (or absence of sensation) which will convince a person that complete dissolution of the body not only possible but an entirely natural process.

\begin{center}
  \Huge{Virtual Reality}\\
  \Huge{and}\\
  \Huge{The Matrix Paradox}
\end{center}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/michael-abrash-senses.png}
  \caption{Traditional 5-sense brain mapping.}
  \label{fig:michael-abrash-senses}
\end{figure}

This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at the Facebook F8 Summit in 2015. \cite{abrashvr} I'm using this slide because it's nicely laid out and because it is incomplete.

Abrash discusses the nature of the ``Matrix Paradox'' \cite{simulationhypothesis} as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. The following might be a slightly better model of reality:

Though a classic sense diagram such as Mr. Abrash’s might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain. Touch is strictly an external sense dealing with skin. These are internal senses.

@@

On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience. On the topic of lacking proprioception: ``We have words like `deaf' and `blind'. We do not have words for being deaf or blind to one's own body.'' In the Matrix, perhaps this is either a bug or a feature flag.

Now that we have our diagram for visualizing the use of our senses, we can address them in terms of various altered mental states. Before we get there, however, we will try to come up with a meaningful visual definition of ``attention'' or ``awareness'' with respect to our senses.

@@

The funny thing about Vipassana is that you are doing it already... you're just terrible at it. Vipassana is the inspection of one's own nervous system with narrowly-focused attention but the precision of this focus is not prescribed by the instruction given for the technique. ``As narrow as possible'', perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip. Your attention is, in part, already on your nervous system, with some degree of focus. To the extent that it is, you are doing Vipassana.

@@

On the first day, your attention might look something like this:
…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions. By the fourth day, your attention will often take a shape like this instead:

@@

If you can’t feel pain in your knees despite the fact that the pain is ``real'', you have effectively eliminated (some) sensory input. This is possible for all sensory input. The 5 classic sense-doors will quite clearly shut down in deep meditative states. Just because your eyes are closed does not make you blind. Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness. This blob which I ``see'' disappears in deep meditative states. Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
Blue Bars:
The New Distractions

@@

These ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar. These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana. Everything in blue, unfortunately, falls into the ``explaining the concept of `white' to a birth-blind man'' category of unexplainable sensory input.

@@

This statement is meant operationally, not existentially. We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event — every operation — occurring in the nervous system has a corresponding event in the brain, and so forth. This mutually recursive process will eventually wear itself out. If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives. However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is ``argh, I hate pain'', which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.

@@

Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body. From ``Where Am I?'' \cite{whereami}:

``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''

``Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''

@@

There is a sketch in ``Where Am I?'' \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:

``Honey, don’t forget what the half-life is on the autonomic nervous system.''

@@

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide is from his 2013 talk ``Performance Testing Java Applications'' \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.

@@

\begin{thebibliography}{9}
\bibitem{thenewsroom}
  Aaron Sorkin.
  \url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
  \textit{The Newsroom}.
  HBO, 2012.

\bibitem{experiencethetruth}
  S.N. Goenka.
  \url{https://www.youtube.com/watch?v=mcbGbN8pFyk&list=PLlb2aTuor6J9j8hACeX6GUmd4WF7QJsbu&index=2}
  \textit{Experience the Truth}.
  Vipassana Research Institute, 1992.

\bibitem{hackerhowto}
  Eric S. Raymond: How To Become A Hacker,
  \url{http://www.catb.org/esr/faqs/hacker-howto.html}

\bibitem{abrashvr}
  Michael Abrash.
  \url{https://www.youtube.com/watch?v=UDu-cnXI8E8}
  \textit{Why Virtual Reality Will Matter to You}.
  Facebook F8 Summit, Fort Mason, 2015.

\bibitem{simulationhypothesis}
  \url{https://en.wikipedia.org/wiki/Simulation_hypothesis}
  \textit{Simulation Hypothesis}.
  Wikipedia.

\bibitem{ramachandranbrainfunction}
  V. S. Ramachandran, Eric L. Altschuler.
  \url{https://academic.oup.com/brain/article/132/7/1693/328686}
  \textit{The use of visual feedback in restoring brain function}.
  Brain, 132(7):1693–1710, July 2009.

\bibitem{proprioception}
  \url{https://en.wikipedia.org/wiki/Proprioception}
  \textit{Proprioception}.
  Wikipedia.

\bibitem{nocioception}
  \url{https://en.wikipedia.org/wiki/Nociception}
  \textit{Nocioception}.
  Wikipedia.

\bibitem{fmrihunger}
  Dagmar Führer, Stefan Zysset, Michael Stumvoll.
  \url{https://www.ncbi.nlm.nih.gov/pubmed/18292747}
  \textit{Brain activity in hunger and satiety: an exploratory visually stimulated fMRI study}
  Obesity, 16(5):945-50, May 2008.

\bibitem{whereami}
  Jad Abumrad, Robert Krulwich.
  \url{http://www.radiolab.org/story/91524-where-am-i/}
  \textit{Where Am I?}
  RadioLab, May 2006.

\bibitem{testingjava}
  Martin Thompson.
  \url{https://www.youtube.com/watch?v=1DuMvpwWHH4}
  \textit{Performance Testing Java Applications}.
  Devoxx UK, 2013.

\end{thebibliography}
